test_types

### 2024-04-22 Sections/test_types.qmd
add docx version; fix igas figure text; re-render everything.

---
title: Full Dissertation2
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
#date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
# page-layout: full
# code-fold: true
# code-tools: true
# toc: true
# toc-location: body
# toc-title: Contents
#cache: true
execute: 
  warning: false
  eval: false
  include: false
format:
 # html: default
  # hugo-md:
  #   echo: false
  #   #include: false
  #   html-math-method: mathjax
  #   output-file: diss_full_hugo22.md
  # gfm:
  #   echo: false
  #   output: false
  #   output-file: diss_full_gfm22.md
  # hikmah-manuscript-pdf:
  #   echo: false
  #   output-file: "tg_full_disser22.pdf"
  #   mainfont: "Linux Libertine O"
  #   mainfontoptions:
  #     - "Numbers=Proportional"
  #     - "Numbers=OldStyle"
  #   mathfont: "Libertinus Math"
  # hikmah-manuscript-docx: 
  #   output-file: "manuscript-ms2.docx"
  #   echo: false
  #   #prefer-html: true
  #   toc: false
  docx:
    echo: false
    prefer-html: true
    output-file: "TG_Full_Dissertation3.docx"
    toc: true
prefer-html: true
---

# Abstract

The past century of research on human learning has produced ample evidence that although learners can improve at almost any task, such improvements are often specific to the trained task, with unreliable or even nonexistent transfer or generalization to novel tasks or conditions [@barnettWhenWhereWe2002; @dettermanCaseProsecutionTransfer1993]. Such generalization challenges are of noteworthy practical relevance, given that educators, trainers, and rehabilitators typically intend for their students to be able to apply what they have learned to new situations. It is therefore important to better understand the factors that influence generalization, and to develop cognitive models that can predict when generalization is likely to occur. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work - the variability of training examples.

Of course, the relationship between training variability and transfer is unlikely to be a simple function wherein increased variation is always beneficial. Numerous studies have found null, or in some cases negative effects of training variation [@deloshExtrapolationSineQua1997; @sinkeviciuteRoleInputVariability2019; @wrisbergVariabilityPracticeHypothesis1987; @vanrossumSchmidtSchemaTheory1990], and many more have suggested that the benefits of variability may depend on additional factors such as prior task experience, the order of training trials, or the type of transfer being measured [@bernikerEffectsTrainingBreadth2014; @braithwaiteEffectsVariationPrior2015; @hahnEffectsCategoryDiversity2005; @lavanEffectsHighVariability2019; @northEffectConsistentVaried2019; @sadakataIndividualAptitudeMandarin2014; @zamanPerceptualVariabilityImplications2021].

In an example of a more complex influence of training variation, [@braithwaiteEffectsVariationPrior2015] trained participants on example problems involving the concept of sampling with replacement (SWR). Training consisted of examples that were either highly similar in their semantic context (e.g. all involving people selecting objects) or in which the surface features were varied between examples (e.g. people choosing objects AND objects selected in a sequence). The experimenters also surveyed how much prior knowledge each participant had with SWR. They found that whether variation was beneficial depended on the prior knowledge of the participants -- such that participants with some prior knowledge benefited from varied training, whereas participants with minimal prior knowledge performed better after training with similar examples. The authors hypothesized that in order to benefit from varied examples, participants must be able to detect the structure common to the diverse examples, and that participants with prior knowledge are more likely to be sensitive to such structure, and thus to benefit from varied training. To test this hypothesis more directly, the authors conducted a 2nd experiment, wherein they controlled prior knowledge by exposing some subjects to a short graphical or verbal pre-training lesson, designed to increase sensitivity to the training examples. Consistent with their hypothesis, participants exposed to the structural sensitivity pre-training benefited more from varied training than the controls participants who benefited more from training with similar examples. Interactions between prior experience and the influence of varied training have also been observed in sensorimotor learning [@guadagnoliRelationshipContextualInterference1999;  @delreyEffectsContextualInterference1982].  @delreyEffectsContextualInterference1982 recruited participants who self-reported either extensive, or very little experience with athletic activities, and then trained participants on a coincident timing task under with either a single constant training velocity, with one of several varied training procedures. Unsurprisingly, athlete participants had superior performance during training, regardless of  condition, and training performance was superior for all subjects in the constant group. Of greater interest is the pattern of testing results from novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received variable training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training.

### The current work

The overarching purpose of this dissertation is to investigate the effects of training variability on learning and generalization within visuomotor skill learning and function learning. Our investigations is structured into two main projects,  each employing distinct experimental paradigms and computational modeling frameworks to elucidate how and when variability in training enhances or impedes subsequent generalization.

In Project 1, we investigated the influence of varied practice in a simple visuomotor projectile launching task. Experiments 1 and 2 compared the performance of constant and varied training groups to assess potential benefits of variability on transfer to novel testing conditions. To account for the observed empirical effects, we introduced the Instance-based Generalization with Adaptive Similarity (IGAS) model. IGAS provides a novel computational approach for quantifying the similarity between training experiences and transfer conditions, while also allowing for variability to influence the generalization gradient itself.

Project 2 shifted focus to the domain of function learning by employing a visuomotor extrapolation task. Across three experiments, we examined how constant and varied training regimes affected learning, discrimination between stimuli, and the ability to extrapolate to novel regions of the function's input space. To model human performance in this task, we fit the influential Associative Learning Model (ALM) and the Extrapolation-Association Model (EXAM) to individual participant data using advanced Bayesian parameter estimation techniques.

{{< pagebreak >}}

# Project 1

```{r}
#| cache: false

pacman::p_load(dplyr,purrr,tidyr,tibble,ggplot2,
  brms,tidybayes, rstanarm,emmeans,broom,bayestestR,
  stringr, here,conflicted, patchwork, knitr, cowplot, RColorBrewer,data.table,gt,
  ggh4x,htmltools,future,furrr,ez)
#options(brms.backend="cmdstanr",mc.cores=4)
#options(digits=2, scipen=999, dplyr.summarise.inform=FALSE, kableExtra.auto_format=FALSE)
walk(c("brms","dplyr","bayestestR","here"), conflict_prefer_all, quiet = TRUE)
walk(c("Display_Functions","fun_alm","fun_indv_fit","fun_model", "prep_model_data","org_functions"), ~source(here::here(paste0("Functions/", .x, ".R"))))
#conflicts_prefer(tidybayes::rstudent_t)

# pacman::p_load(dplyr,purrr,tidyr,ggplot2, data.table, here, patchwork, conflicted, 
#                stringr,future,furrr, knitr, reactable,ggstance, htmltools,
#                ggdist,ggh4x,brms,tidybayes,emmeans,bayestestR, gt)

#pacman::p_load(papaja, tinytex, RColorBrewer, kableExtra, cowplot)
source(here::here('Functions/IGAS_ProcessFunctions.R'))

#options(brms.backend="cmdstanr",mc.cores=4)
initial_contrasts <- options("contrasts")$contrasts
options(digits=2, scipen=999, dplyr.summarise.inform=FALSE, contrasts = c("contr.sum", "contr.poly"))
defaultContrasts = options()$contrasts

library(tidybayes)

theme_set(theme_classic())
# load the processed data from experiment 1 and 2
e1 <- readRDS(here::here("data/igas_e1_cleanedData-final.rds")) %>% mutate(initialVelocityX=X_Velocity,initialVelocityY=Y_Velocity,stageInt=as.numeric(as.character(experimentStage)))
e2<- readRDS(here::here('data/igas_e2_cleanedData-final.rds')) %>% mutate(initialVelocityX=X_Velocity,initialVelocityY=Y_Velocity)
# load subject similarity data - computed with the IGAS model in 'IGAS-SimModel.R'
e2_sim <- readRDS(here::here('data/IGAS_Similarity-Performance.rds'))

# options(contrasts = c("contr.sum", "contr.poly"), kableExtra.auto_format=FALSE)
# defaultContrasts = options()$contrasts
# theme_set(theme_classic())

dodge <- position_dodge(width = 0.9)
e2GrpPos <- c("400","500","625","675","800","900")
e2Grp <- paste("Constant","Constant", "Constant","Constant","Constant","Constant", "Varied")
e2Labels <- paste(c("400\n Constant","500\n Constant","625\n Constant","675\n Constant",
                   "800\n Constant","900\n Constant","500-800\n Varied"),sep="")

e1Pos <- c("610","760","835","910")
e1Var <- paste("Varied Train Position","Constant Train Position", "Novel Position", "Varied Training Position")
e1Labels<- paste(c("610\n Varied Trained","760\n Constant Trained","835\n Novel Location","910\n Varied Trained"),sep="")
```

# Abstract

Exposing learners to variability during training has been demonstrated to improve performance in subsequent transfer testing. Such variability benefits are often accounted for by assuming that learners are developing some general task schema or structure. However much of this research has neglected to account for differences in similarity between varied and constant training conditions. In a between-groups manipulation, we trained participants on a simple projectile launching task, with either varied or constant conditions. We replicate previous findings showing a transfer advantage of varied over constant training. Furthermore, we show that a standard similarity model is insufficient to account for the benefits of variation, but, if the model is adjusted to assume that varied learners are tuned towards a broader generalization gradient, then a similarity-based model is sufficient to explain the observed benefits of variation. Our results therefore suggest that some variability benefits can be accommodated within instance-based models without positing the learning of some schemata or structure.

@@

e2$stage <- factor(e2$stage, levels = c("Beginning", "Middle", "End","Transfer"),ordered = TRUE)

exp2TrainPosition <- e2  %>% filter(stage!="Transfer") %>%ungroup() %>% 
  group_by(sbjCode,Group2,conditType,trainHalf,positionX) %>% 
  summarise(MeanTargetDistance=mean(AbsDistFromCenter))%>% as.data.frame()

exp2TrainPosition3 <- e2  %>% filter(stage!="Transfer") %>%ungroup() %>% 
  mutate(globalAvg=mean(AbsDistFromCenter),globalSd=sd(AbsDistFromCenter)) %>% 
  group_by(sbjCode,positionX) %>% mutate(scaledDev = scaleVar(globalAvg,globalSd,AbsDistFromCenter)) %>%ungroup() %>%
  group_by(sbjCode,Group2,conditType,stage,positionX) %>% 
  summarise(MeanTargetDistance=mean(AbsDistFromCenter),MeanScaledDev=mean(scaledDev,trim=.05))%>% as.data.frame()

exp2Train <- e2  %>% filter(stage!="Transfer")  %>% 
  group_by(sbjCode,Group2,conditType,trainHalf) %>% 
  summarise(MeanTargetDistance=mean(AbsDistFromCenter)) %>% as.data.frame()

exp2Train3 <- e2  %>% filter(stage!="Transfer")  %>% ungroup() %>% 
  mutate(globalAvg=mean(AbsDistFromCenter),globalSd=sd(AbsDistFromCenter)) %>% 
  group_by(sbjCode,positionX) %>% mutate(scaledDev = scaleVar(globalAvg,globalSd,AbsDistFromCenter)) %>%ungroup() %>%
  group_by(sbjCode,Group2,conditType,stage) %>% 
  summarise(MeanTargetDistance=mean(AbsDistFromCenter),MeanScaledDev=mean(scaledDev,trim=.05)) %>% as.data.frame()

transfer <- filter(e2, stage=="Transfer") %>% droplevels() %>% select(-trainHalf,-initialVelocityY,ThrowPosition2)%>% ungroup()
transfer <- transfer %>% group_by(positionX) %>% mutate(globalAvg=mean(AbsDistFromCenter),globalSd=sd(AbsDistFromCenter)) %>% 
  group_by(sbjCode,positionX) %>% mutate(scaledDev = scaleVar(globalAvg,globalSd,AbsDistFromCenter)) %>%ungroup()

transfer <- transfer %>% group_by(sbjCode,positionX) %>% mutate(ind=1,testPosIndex=cumsum(ind),posN=max(testPosIndex)) %>%
  select(-ind) %>% mutate(testHalf = case_when(testPosIndex<15 ~"1st Half",testPosIndex>=15 ~"2nd Half")) %>% rstatix::convert_as_factor(testHalf)

variedTest <- transfer %>% filter(condit==7) %>% mutate(extrapolate=ifelse(positionX=="900" | positionX=="400","extrapolation","interpolation")) 
constantTest <- transfer %>% filter(condit!=7) %>% mutate(extrapolate=ifelse(distFromTrain==0,"interpolation","extrapolation"))

transfer <- rbind(variedTest,constantTest)
transfer<- transfer %>% mutate(novel=ifelse(distFromTrain3==0,"trainedLocation","novelLocation"))%>% rstatix::convert_as_factor(novel,extrapolate)

transfer <- transfer %>% relocate(sbjCode,condit2,Group,conditType2,stage,trial,novel,extrapolate,positionX,AbsDistFromCenter,globalAvg,globalSd,scaledDev,distFromTrain3) %>% ungroup()

# novelAll <- transfer %>% filter(distFromTrain!=0, distFromTrain3!=0) %>% select(-globalAvg,-globalSd,-scaledDev)%>% droplevels() %>% ungroup()
# novelAll <- novelAll %>% group_by(positionX) %>%
#  mutate(globalAvg=mean(AbsDistFromCenter),globalSd=sd(AbsDistFromCenter)) %>% 
#   group_by(sbjCode,positionX) %>% mutate(scaledDev = scaleVar(globalAvg,globalSd,AbsDistFromCenter)) %>%ungroup()

novelAll <- transfer %>% filter(distFromTrain!=0, distFromTrain3!=0)
novelAllMatched <- novelAll %>% filter(condit!=5,condit!=2)

constantIden <- transfer %>% filter(condit !=7,distFromTrain==0) # only constant groups from their training position
variedTest <- transfer %>% filter(condit==7) # only varied testing
variedVsIden <- rbind(constantIden,variedTest) # all varied combined with constant identity

variedNovel <- variedTest %>% filter(distFromTrain3 !=0) # removes 500 and 800 from varied
constantIden2 <- transfer %>% filter(condit !=7,condit!=5,condit!=2,distFromTrain==0) # only constant groups from training position 400,625,675,900
variedVsNovelIden <- rbind(constantIden2,variedNovel) # novel positions for varied, trained for constant

exp2.Test <- transfer %>%group_by(sbjCode,conditType,positionX,ThrowPosition)%>%
  summarise(MeanTargetDeviance = mean(AbsDistFromCenter,trim=.05),MeanScaledDev=mean(scaledDev,trim=.05)) %>%ungroup() %>% as.data.frame()

exp2.Test2 <- exp2.Test %>% group_by(sbjCode,conditType)%>%
  summarise(MeanTargetDeviance = mean(MeanTargetDeviance),MeanScaledDev=mean(MeanScaledDev)) %>%ungroup() %>% as.data.frame()

exp2.Test7 <- transfer %>%group_by(Group2,sbjCode,positionX,Group,conditType,ThrowPosition4) %>% 
  summarise(MeanTargetDeviance = mean(AbsDistFromCenter,trim=.05),MeanScaledDev=mean(scaledDev,trim=.05)) %>% as.data.frame()

exp2.Test7.agg <- exp2.Test7  %>%group_by(Group2,sbjCode,Group,conditType) %>% 
  summarise(MeanTargetDeviance = mean(MeanTargetDeviance),MeanScaledDev=mean(MeanScaledDev)) %>% as.data.frame()

exp2.Test7.agg2 <- exp2.Test7  %>%group_by(sbjCode,conditType) %>% 
  summarise(MeanTargetDeviance = mean(MeanTargetDeviance),MeanScaledDev=mean(MeanScaledDev)) %>% as.data.frame()

@@

```{r}
#| label: fig-e2test1
#| fig.cap: A comparison of throwing location that are identical to those trained by the constant participants (e.g., constant participants trained at position 900, tested from position 900), which are also novel to the varied-trained participants (thus excluding positions 500 and 800). Error bars indicate standard error of the mean.

sum.variedVsNovelIden <- variedVsNovelIden  %>%
  group_by(sbjCode,conditType,positionX) %>% 
  summarise(MeanTargetDev=mean(AbsDistFromCenter,trim=.05),MeanScaledDev=mean(scaledDev,trim=.05),.groups="keep") %>% as.data.frame()

e2Test2 <- sum.variedVsNovelIden %>% ggplot(aes(x=positionX,y=MeanTargetDev,group=conditType,fill=conditType))+geom_bar(stat="summary",position=dodge,fun="mean")+ stat_summary(fun.data=mean_se,geom="errorbar",position=dodge,width=.5)+ylab("Mean Distance From Center Of Target") +xlab("Testing Location")+theme(plot.title = element_text(hjust = 0.5))+guides(fill=guide_legend(title="Training Condition"))+theme(legend.title.align=.25)

e2Test2

\
\

```{r}
#| label: tbl-e2tab5
#| tbl-cap:  Testing performance from the locations trained by constant participants and novel to varied participants. Locations 500 and 800 are not included as these were trained by the varied participants. Table presents Mean absolute deviation from the center of the target, and standard deviations in parenthesis.

tab4=sum.variedVsNovelIden %>% rename(Condition="conditType") %>% group_by(Condition,positionX) %>%
  summarise(Mean=round(mean(MeanTargetDev),2),sd=round(sd(MeanTargetDev),2),.groups="keep")

tab4=tab4 %>% group_by(Condition) %>% 
   mutate(GroupAvg=round(mean(Mean),2),groupSd=round(sd(Mean),2))

tab4 = tab4 %>% mutate(msd=paste(Mean,"(",sd,")",sep=""),gsd=paste(GroupAvg,"(",groupSd,")",sep="")) %>% select(positionX,Condition,msd,gsd)%>%pivot_wider(names_from = Condition,values_from=c(msd,gsd))
 tab4=tab4[,1:3]

kable(tab4,escape=FALSE,booktabs=TRUE,col.names=c("Position","Constant","Varied"),align=c("l")) # %>% kableExtra::kable_styling(position="left") #%>% kable_classic() #%>% footnote(general=captionText,general_title = "")

### Experiment 2 Discussion

The results of Experiment 2 largely conform to the findings of Experiment 1. Participants in both varied and constant conditions improved at the task during the training phase. We did not observe the common finding of training under varied conditions producing worse performance during acquisition than training under constant conditions [@catalanoDistantTransferCoincident1984a; @wrisbergVariabilityPracticeHypothesis1987], which has been suggested to relate to the subsequent benefits of varied training in retention and generalization testing [@soderstromLearningPerformanceIntegrative2015]. However our finding of no difference in training performance between constant and varied groups has been observed in previous work [@chuaPracticeVariabilityPromotes2019; @moxleySchemaVariabilityPractice1979; @pigottMotorSchemaStructure1984].

In the testing phase, our varied group significantly outperformed the constant conditions in both a general comparison, and in an analysis limited to novel throwing positions. The observed benefit of varied over constant training echoes the findings of many previous visuomotor skill learning studies that have continued to emerge since the introduction of Schmidt's influential Schema Theory [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @goodwinEffectDifferentQuantities1998; @mccrackenTestSchemaTheory1977; @moxleySchemaVariabilityPractice1979; @newellVariabilityPracticeTransfer1976; @pigottMotorSchemaStructure1984; @rollerVariablePracticeLenses2001; @schmidtSchemaTheoryDiscrete1975; @willeyLongtermMotorLearning2018; @wrisbergVariabilityPracticeHypothesis1987; @wulfEffectTypePractice1991]. We also join a much smaller set of research to observe this pattern in a computerized task [@seowTransferEffectsVaried2019]. One departure from the experiment 1 findings concerns the pattern wherein the varied group outperformed the constant group even from the training position of the constant group, which was significant in experiment 1, but did not reach significance in experiment 2. Although this pattern has been observed elsewhere in the literature [@goodeSuperiorityVariableRepeated2008; @kerrSpecificVariedPractice1978], the overall evidence for this effect appears to be far weaker than for the more general benefit of varied training in conditions novel to all training groups.

## Computational Model

*Controlling for the similarity between training and testing.* The primary goal of Experiment 2 was to examine whether the benefits of variability would persist after accounting for individual differences in the similarity between trained and tested throwing locations. To this end, we modelled each throw as a two-dimensional point in the space of x and y velocities applied to the projectile at the moment of release. For each participant, we took each individual training throw, and computed the similarity between that throw and the entire population of throws within the solution space for each of the 6 testing positions. We defined the solution space empirically as the set of all combinations of x and y throw velocities that resulted in hitting the target. We then summed each of the trial-level similarities to produce a single similarity for each testing position score relating how the participant threw the ball during training and the solutions that would result in target hits from each of the six testing positions -- thus resulting in six separate similarity scores for each participant. @fig-taskSpace visualizes the solution space for each location and illustrates how different combinations of x and y velocity result in successfully striking the target from different launching positions. As illustrated in @fig-taskSpace, the solution throws represent just a small fraction of the entire space of velocity combinations used by participants throughout the experiment.

```{r}
#| label: fig-taskSpace
#| fig.cap: A) A visual representation of the combinations of throw parameters (x and y velocities applied to the ball at launch), which resulted in target hits during the testing phase. This empirical solution space was compiled from all of the participants in Experiment 2. B) shows the solution space within the context of all of the throws made throughout the testing phase of the experiment.
#| fig-width: 11
#| fig-height: 9
#| 
taskspace <- e2 %>% filter(AbsDistFromCenter<900)
taskspace$hitOrMiss <- ifelse(taskspace$trialType==11,"Hit Target","Missed Target")

solSpace <- e2 %>% filter(trialType==11)
#solSpace %>% ggplot(aes(x=X_Velocity,y=Y_Velocity)) + geom_point(aes(colour=ThrowPosition),alpha=0.58) + ggtitle("")

solSpace$Result = ifelse(solSpace$ThrowPosition==400,"400",solSpace$ThrowPosition)
solSpace$Result = ifelse(solSpace$ThrowPosition==500,"500",solSpace$Result)
solSpace$Result= ifelse(solSpace$ThrowPosition==625,"625",solSpace$Result)
solSpace$Result = ifelse(solSpace$ThrowPosition==675,"675",solSpace$Result)
solSpace$Result = ifelse(solSpace$ThrowPosition==800,"800",solSpace$Result)
solSpace$Result = ifelse(solSpace$ThrowPosition==900,"900",solSpace$Result)

missSpace <- e2 %>% filter(trialType !=11)
missSpace$Result = "Missed Target"
solSpace$Result <- solSpace$Result

# the usual method of changing the legend title does not seem to work after the colours are manually scaled. 
# multiplied velocoties by -1 to make the axes less confusing
ss=solSpace %>% ggplot(aes(x=X_Velocity*-1,y=Y_Velocity*-1)) + 
  geom_point(aes(colour=Result),alpha=0.6) + 
  scale_color_manual(values =brewer.pal(n=6,name="Set1"))+
  labs(colour="Target Hit Thrown from Position:") + xlab("X Release Velocity") + ylab("Y Release Velocity")+ggtitle("A")

fullSpace <- rbind(missSpace,solSpace)

fs<- fullSpace %>% ggplot(aes(x=X_Velocity*-1,y=Y_Velocity*-1,colour=Result)) + 
  geom_point(aes(),alpha=0.6) + scale_color_manual(values =brewer.pal(n=7,name="Set1"))+
  labs(colour="Target Hit or Miss From Position:") + xlab("X Release Velocity") + ylab("Y Release Velocity") +ggtitle("B")

library(patchwork)
ss/fs

For each individual trial, the Euclidean distance (Equation 1) was computed between the velocity components (x and y) of that trial and the velocity components of each individual solution throw for each of the 6 positions from which participants would be tested in the final phase of the study. The P parameter in Equation 1 is set equal to 2, reflecting a Gaussian similarity gradient. Then, as per an instance-based model of similarity [@loganInstanceTheoryAttention2002a; @nosofskySimilarityScalingCognitive1992], these distances were multiplied by a sensitivity parameter, c, and then exponentiated to yield a similarity value. The parameter c controls the rate with which similarity-based generalization drops off as the Euclidean distance between two throws in x- and y-velocity space increases. If c has a large value, then even a small difference between two throws' velocities greatly decreases the extent of generalization from one to the other. A small value for c produces broad generalization from one throw to another despite relatively large differences in their velocities. The similarity values for each training individual throw made by a given participant were then summed to yield a final similarity score, with a separate score computed for each of the 6 testing positions. The final similarity score is construable as index of how accurate the throws a participant made during the training phase would be for each of the testing positions.

**Equation 1:** $$ Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}}) $$

**Equation 2:**
$$ d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 } $$

A simple linear regression revealed that these similarity scores were significantly predictive of performance in the transfer stage, t =-15.88, p\<.01, $r^2$=.17, such that greater similarity between training throws and solution spaces for each of the test locations resulted in better performance. We then repeated the group comparisons above while including similarity as a covariate in the model. Comparing the varied and constant groups in testing performance from all testing positions yielded a significant effect of similarity, F(1, 205)=85.66, p\<.001, $\eta^{2}_G$ =.29, and also a significant effect of condition (varied vs. constant), F(1, 205)=6.03, p=.015, $\eta^{2}_G$ =.03. The group comparison limited to only novel locations for the varied group pit against trained location for the constant group resulted in a significant effect of similarity, F(1,148)=31.12, p\<.001, $\eta^{2}_G$ =.18 as well as for condition F(1,148)=11.55, p\<.001, $\eta^{2}_G$ =.07. For all comparisons, the pattern of results was consistent with the initial findings from Experiment 2, with the varied group still performing significantly better than the constant group.

### Fitting model parameters separately by group

To directly control for similarity in Experiment 2, we developed a model-based measure of the similarity between training throws and testing conditions. This similarity measure was a significant predictor of testing performance, e.g., participants whose training throws were more similar to throws that resulted in target hits from the testing positions, tended to perform better during the testing phase. Importantly, the similarity measure did not explain away the group-level benefits of varied training, which remained significant in our linear model predicting testing performance after similarity was added to the model. However, previous research has suggested that participants may differ in their level of generalization as a function of prior experience, and that such differences in generalization gradients can be captured by fitting the generalization parameter of an instance-based model separately to each group [@hahnEffectsCategoryDiversity2005; @lambertsFlexibleTuningSimilarity1994]. Relatedly, the influential Bayesian generalization model developed by @tenenbaumGeneralizationSimilarityBayesian2001a predicts that the breadth of generalization will increase when a rational agent encounters a wider variety of examples. Following these leads, we assume that in addition to learning the task itself, participants are also adjusting how generalizable their experience should be. Varied versus constant participants may be expected to learn to generalize their experience to different degrees. To accommodate this difference, the generalization parameter of the instance-based model (in the present case, the c parameter) can be allowed to vary between the two groups to reflect the tendency of learners to adaptively tune the extent of their generalization. One specific hypothesis is that people adaptively set a value of c to fit the variability of their training experience [@nosofskyExemplarbasedAccountsMultiplesystem2000; @sakamotoTrackingVariabilityLearning2006]. If one's training experience is relatively variable, as with the variable training condition, then one might infer that future test situations will also be variable, in which case a low value of c will allow better generalization because generalization will drop off slowly with training-to-testing distance. Conversely, if one's training experience has little variability, as found in the constant training conditions, then one might adopt a high value of c so that generalization falls off rapidly away from the trained positions.

To address this possibility, we compared the original instance-based model of similarity fit against a modified model which separately fits the generalization parameter, c, to varied and constant participants. To perform this parameter fitting, we used the optim function in R, and fit the model to find the c value(s) that maximized the correlation between similarity and testing performance.

Both models generate distinct similarity values between training and testing locations. Much like the analyses in Experiment 2, these similarity values are regressed against testing performance in models of the form shown below. As was the case previously, testing performance is defined as the mean absolute distance from the center of the target (with a separate score for each participant, from each position).

Linear models 1 and 3 both show that similarity is a significant predictor of testing performance (p\<.01). Of greater interest is the difference between linear model 2, in which similarity is computed from a single c value fit from all participants (Similarity1c), with linear model 4, which fits the c parameter separately between groups (Similarity2c). In linear model 2, the effect of training group remains significant when controlling for Similarity1c (p\<.01), with the varied group still performing significantly better. However, in linear model 4 the addition of the Similarity2c predictor results in the effect of training group becoming nonsignificant (p=.40), suggesting that the effect of varied vs. constant training is accounted for by the Similarity2c predictor. Next, to further establish a difference between the models, we performed nested model comparisons using ANOVA, to see if the addition of the training group parameter led to a significant improvement in model performance. In the first comparison, ANOVA(Linear Model 1, Linear Model 2), the addition of the training group predictor significantly improved the performance of the model (F=22.07, p\<.01). However, in the second model comparison, ANOVA (Linear model 3, Linear Model 4) found no improvement in model performance with the addition of the training group predictor (F=1.61, p=.20).

Finally, we sought to confirm that similarity values generated from the adjusted Similarity2c model had more predictive power than those generated from the original Similarity1c model. Using the BIC function in R, we compared BIC values between linear model 1 (BIC=14604.00) and linear model 3 (BIC = 14587.64). The lower BIC value of model 3 suggests a modest advantage for predicting performance using a similarity measure computed with two c values over similarity computed with a single c value. When fit with separate c values, the best fitting c parameters for the model consistently optimized such that the c value for the varied group (c=.00008) was smaller in magnitude than the c value for the constant group(c= .00011). Recall that similarity decreases as a Gaussian function of distance (equation 1 above), and a smaller value of c will result in a more gradual drop-off in similarity as the distance between training throws and testing solutions increases.

In summary, our modeling suggests that an instance-based model which assumes equivalent generalization gradients between constant and varied trained participants is unable to account for the extent of benefits of varied over constant training observed at testing. The evidence for this in the comparative model fits is that when a varied/constant dummy-coded variable for condition is explicitly added to the model, the variable adds a significant contribution to the prediction of test performance, with the variable condition yielding better performance than the constant conditions. However, if the instance-based generalization model is modified to assume that the training groups can differ in the steepness of their generalization gradient, by incorporating a separate generalization parameter for each group, then the instance-based model can account for our experimental results without explicitly taking training group into account. Henceforth this model will be referred to as the Instance-based Generalization with Adaptive Similarity (IGAS) model.

## Project 1 General Discussion

Across two experiments, we found evidence in support of the benefits of
variability hypothesis in a simple, computerized projectile throwing task. Generalization was observed in both constant and varied participants, in that both groups tended to perform better at novel positions in the testing phase than did participants who started with those positions in the training phase. However, varied trained participants consistently performed better than constant trained participants, in terms of both the testing phase in general, and in a comparison that only included untrained positions. We also found some evidence for the less commonly observed pattern wherein varied-trained participants outperform constant-trained participants even from conditions identical to the constant group training [@goodeSuperiorityVariableRepeated2008; @greenPracticeVariabilityTransfer1995a; @kerrSpecificVariedPractice1978]. In Experiment 1 varied participants performed significantly better on this identity comparison. In Experiment 2, the comparison was not significant initially, but became significant after controlling for the similarity measure that incorporates only a single value for the steepness of similarity-based generalization (c). Furthermore, we showed that the general pattern of results from Experiment 2 could be parsimoniously accommodated by an instance-based similarity model, but only with the assumption that constant and varied participants generalize their training experience to different degrees. Our results thus suggest that the benefits of variation cannot be explained by the varied-trained participants simply covering a broader range of the task space. Rather, the modeling suggests that varied participants also learn to adaptively tune their generalization function such that throwing locations generalize more broadly to one another than they do in the constant condition. A learning system could end up adopting a higher c value in the constant than variable training conditions by monitoring the trial-by-trial variability of the training items. The c parameter would be adapted downwards when adjacent training items are dissimilar to each other and adapted upwards when adjacent training items are the same. In this fashion, contextually appropriate c values could be empirically learned. This learning procedure would capture the insight that if a situation has a high amount variability, then the learner should be predisposed toward thinking that subsequent test items will also show considerable variability, in which case generalization gradients should be broad, as is achieved by low values for c.

Also of interest is whether the IGAS model can predict the pattern of results wherein the varied condition outperforms the constant condition even from the position on which the constant condition trained. Although our models were fit using all of the Experiment 2 training and testing data, not just that of the identity comparisons, in @fig-Toy-Model-dis we demonstrate how a simplified version of the IGAS model could in principle produce such a pattern. In addition to the assumption of differential generalization between varied and constant conditions, our simplified model makes explicit an assumption that is incorporated into the full IGAS model -- namely that even when being tested from a position identical to that which was trained, there are always some psychological contextual differences between training and testing throws, resulting in a non-zero dissimilarity.

>>>>

---
title: "test_types"
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
#toc-title: Contents
toc-depth: 4
cache: false
#format-links: [hikmah-manuscript-pdf, gfm]
execute: 
  warning: false
  eval: false
format:
#   html: default
#   hugo-md:
#     echo: false
#     include: true
#     html-math-method: mathjax
#     output-file: test_hugo.md
#   gfm:
#     echo: false
#     output: true
#     output-file: testl_gfm.md
#   hikmah-manuscript-pdf:
#     echo: false
#     output-file: "test.pdf"
#     mainfont: "Linux Libertine O"
#     mainfontoptions:
#       - "Numbers=Proportional"
#       - "Numbers=OldStyle"
#     mathfont: "Libertinus Math"
  docx:
    echo: false
    prefer-html: true
    output-file: "teset.docx"
    toc: true
prefer-html: true
---

**Approximate Bayesian Computation**

```{r}
#| echo: false
#| 
pacman::p_load(dplyr,purrr,tidyr,tibble,ggplot2,
  brms,tidybayes, rstanarm,emmeans,broom,bayestestR,
  stringr, here,conflicted, patchwork, knitr,gt)

out_type <- knitr::opts_knit$get("rmarkdown.pandoc.to")
print(out_type)
primary=out_type == "html" || out_type == "pdf" || out_type =="latex"

fmt_out <- knitr::pandoc_to()
print(fmt_out)

out_type %in% c("html","pdf")
# out_type = "html"
out_type == "html" || out_type == "pdf" || out_type == "docx"

@@

# create examples of fancy looking gt tables using cars data
cars <- cars %>% as_tibble()

# create examples of fancy looking gt tables using flowers data
# create flowers data
flowers <- iris %>% as_tibble()
# flowers
# # A tibble: 150 × 5
#    Sepal.Length Sepal.Width Petal.Length Petal.Width Species
#           <dbl>       <dbl>        <dbl>       <dbl> <fct>  
#  1          5.1         3.5          1.4         0.2 setosa 
#  2          4.9         3            1.4         0.2 setosa 
#  3          4.7         3.2          1.3         0.2 setosa 
#  4          4.6         3.1          1.5         0.2 setosa 
#  5          5           3.6          1.4         0.2 setosa 
#  6          5.4         3.9          1.7         0.4 setosa 
#  7          4.6         3.4          1.4         0.3 setosa

# create examples of fancy looking gt tables using flowers data

if (primary){
flowers |> gt() |> tab_header(
  title = "Iris Flower Data",
  subtitle = "A tibble: 150 × 5"
) |> fmt_number(
  columns = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width),
  decimals = 1) |> 
  tab_spanner(
    label = "Sepal Characteristics",
    columns = c(Sepal.Length, Sepal.Width)
  ) 
} else {
    flowers |> kable() }

@@
